Designing Types for R, Empirically

### 2020-09-11 paper/main.tex
added alternate class tyoe unification discussion

In this paper, we will attempt infer type signature for functions. We now
introduce some relevant concepts.  Functions can be called with named
parameters, arguments can have default values, and R also support variable
argument lists.  To illustrate all of this in a single example, consider:

\begin{lstlisting}
  f <- function(x, ..., y=if(z==0) 1, z=0) x + y + if (missing(...)) 0 else c(...)
\end{lstlisting}

@@

\noindent
The above hints at the polymorphism of the language, \k f, may return a
vector of integer or of double, the length of the vector depends on the
length of the varargs and of \k x and \k y.  The language does not really
differentiate between scalar and vectors.  Some more exotic types that can
be encountered include vectors of complex number and list of arbitrary
types. Function \k f can be invoked with those as well.

@@

R has one builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{rtypes} lists all of the builtin types that are
provided by the language. They are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R. But, in many
context a \k{logical} will convert to \k{integer}, and an \k{integer} will
convert to \k{double}.  Some odd conversions can occur in corner cases, such
as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
vector, as the double is converted to an integer. R does not distinguish
between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}. Finally all
vectorized data types have a distinguished missing value denoted by
\code{NA}. The default type of \code{NA} is \k{logical}. We can see that
\code{typeof(NA)=="logical"}, but NA inhabits every type:
\code{typeof(c(1,NA)[2])=="double"}.

@@

 We observed a number of other interesting patterns in the contract assertion failure data.

@@

\section{Conclusion}

>>>>

In this paper, we will attempt infer type signature for functions. 
We begin by introducing some relevant concepts.

The basic building blocks of R are vectors of primitives.
Characters (strings), integers, logicals (booleans), doubles, complex numbers, and raw bytes make up R's primitive values.
When we say that R is {\it vectorized}, we mean that primitive values are {\it always} vectors.
For example, in R one might see \code{2.5} and think it to be a scalar double, but it is in actuality a unit-length vector containing a single double.
Vectors can also be explicitly constructed using the vector constructor \code{c(...)}, 
For example, \code{c(1L, 2L, 3L)} creates a vector of three integers (in R, integer literals are denoted by an \code{L}).

The main object of our work is R functions, which have a number of quirks.
Arguments to functions in R can be assigned default values (which can be arbitrary expressions!), and R functions can support variable argument lists, and functions can be called by specifying argument names explicitly in the call.
To illustrate all of this in a single example, consider:

\begin{lstlisting}
f <- function(x, ..., y=if(z==0) 1, z=0) { 
  x + y + if (missing(...)) 0 else c(...)
}
\end{lstlisting}

@@

\noindent
The above hints at the polymorphism of the language: \k f may return a
vector of integers or of doubles, the length of the vector depends on the
length of the varargs, and of \k x and \k y.  As we mentioned, the language does not really
differentiate between scalar and vectors.  Some more exotic types that can
be encountered include vectors of complex number and list of arbitrary
types. Function \k f can be invoked with those as well.

@@

R has one builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{rtypes} lists all of the builtin types that are
provided by the language, which are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R, but in many
context a \k{logical} will convert to \k{integer}, and an \k{integer} will
convert to \k{double}.  Some odd conversions can occur in corner cases, such
as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
vector, as the double is converted to an integer. R does not distinguish
between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}. Finally all
vectorized data types have a distinguished missing value denoted by
\code{NA}. The default type of \code{NA} is \k{logical}. We can see that
\code{typeof(NA)=="logical"}, but NA inhabits every type:
\code{typeof(c(1,NA)[2])=="double"}.

@@

 We drew a number of other observations from the contract assertion failure data.

@@

%
%
%
%
\subsection{Discussion}

\AT{Talk about how we thought do to a different unification for class types.}

As a closing point of discussion, we would like to highlight an alternate strategy for unifying class types that we considered.
Recall that in our proposed type annotation language, classes are represented as \CLASS{{\tt A}} for something of class {\tt A}, and \CLASS{{\tt A}, {\tt B}} for something with two classes {\tt A} and {\tt B}.

During the robustness phase of our evaluation, we noted a number of class-based type errors such as a \CLASS{{\tt data.frame}} being passed to an argument expecting, e.g., a \CLASS{{\tt data.frame}, {\tt tbl}, {\tt tbl\_df}}.
This represents something of a limitation of our approach, as data frames and {\tt dplyr} tibbles are related and convertible (you can think of a tibble as a subclass of a data frame)---a fact not represented by our types.
One option to mitigate this would be to include a type such as \CLASS{{\tt data.frame}, ...} that must include {\it at least} the {\tt data.frame} class, but this would not altogether solve this problem.
For one, to generate this type we would either need to a) include some type hierarchy in our type system, which is outside of the scope of this work, or b) somehow infer the type with \typetracer.
To create this type, \typetracer would have to observe values of {\it both} type \CLASS{{\tt data.frame}} and \CLASS{{\tt data.frame}, {\tt tbl}, {\tt tbl\_df}}, which would not solve the aforementioned problem as many {\tt dplyr} functions are only called with tibbles.

Further, the data table, another data frame alternative, while being coercible to and from a data frame, only has \CLASS{{\tt data.table}} (not including {\tt data.frame} as part of its class).
This alternate class type would also not adequately capture this behaviour, and even worse, if a \CLASS{{\tt data.frame}} and \CLASS{{\tt data.table}} were to co-occur in an argument type, we would unify it to \CLASS{...}, which is not very informative.

For these reasons, we believe our unification strategy for class types to be adequate until we extend our type system to handle R's wild and varied object orientation frameworks.

%
%
%
%
%
%
\section{Conclusion}
